---
title: "Parties and After Parties in NYC"
author: "iair kleiman"
output:
  html_document: default
  html_notebook: default
---

## Intro

The idea of this script is to be able to identify the party 
and after party spots in NYC based on the data provided by Uber to fivethirtyeight.

This data provides the pickup coordinates of all the Uber trips during most of 2014.

Based on this, I will cluster the data and identify the most popular spots in NYC over the weekends, in 2 day segments.

1. 12am to 3am (for departing a normal party spot)
2. 3 am to 6am (for departing an afterparty spot)



```{r Package Loading, include=FALSE}
#library(tmap)
library(leaflet)
# library(rMaps)
library(maps)
library(rgdal)
library(sp)
library(RColorBrewer)
library(lubridate)
library(tidyverse)

```

```{r File Loading, message=FALSE, error=FALSE,  include=FALSE}
uber_raw_data_apr14 <- read_csv("uber-raw-data-apr14.csv")
uber_raw_data_may14 <- read_csv("uber-raw-data-may14.csv")
uber_raw_data_jun14 <- read_csv("uber-raw-data-jun14.csv")
uber_raw_data_jul14 <- read_csv("uber-raw-data-jul14.csv")
uber_raw_data_aug14 <- read_csv("uber-raw-data-aug14.csv")
uber_raw_data_sep14 <- read_csv("uber-raw-data-sep14.csv")

```

```{r Polygon Tribble, include=FALSE}
p_manhattan <- tribble(
  ~Lat, ~Lon,
  40.88, -73.93,
  40.753, -74.02,
  40.695, -74.02,
  40.71, -73.97,
  40.74, -73.965,
  40.793, -73.914,
  40.81, -73.93,
  40.835, -73.934,
  40.87, -73.909) %>% 
    mutate(Borough = "Manhattan")

p_bronx <- tribble(
  ~Lat, ~Lon,
  40.793, -73.914,
  40.81, -73.93,
  40.835, -73.934,
  40.87, -73.909,
  40.88, -73.93,
  40.918, -73.92,
  40.875, -73.75,
  40.80, -73.7886) %>% 
    mutate(Borough = "Bronx")

p_queens <-  tribble(
  ~Lat, ~Lon,
  40.74, -73.965,
  40.793, -73.914,
  40.80, -73.788,
  40.75, -73.7,
  40.576, -73.735,
  40.53, -73.967,
  40.558, -73.955,
  40.59, -73.836,
  40.695, -73.87,
  40.68, -73.90,
  40.729, -73.93) %>% 
        mutate(Borough = "Queens")

p_brooklyn <-  tribble(
  ~Lat, ~Lon,
  40.74, -73.965,
  40.71, -73.97,
  40.695, -74.02,
  40.65, -74.05,
  40.55, -74.05,
  40.558, -73.955,
  40.59, -73.836,
  40.695, -73.87,
  40.68, -73.90,
  40.729, -73.93) %>% 
        mutate(Borough = "Brooklyn")

p_staten <-  tribble(
  ~Lat, ~Lon,
  40.65, -74.05,
  40.55, -74.05,
  40.5, -74.26,
  40.645, -74.19) %>% 
        mutate(Borough = "Staten Island")

NY_poly <- bind_rows(p_manhattan, p_bronx, p_queens, p_brooklyn, p_staten) %>%  
    group_by(Borough) %>%
    do(poly=select(., Lat, Lon) %>% Polygon()) %>%
    rowwise() %>%
    do(polys=Polygons(list(.$poly),.$Borough)) %>%
    {SpatialPolygons(.$polys)}

proj4string(NY_poly) = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

df<- data.frame(id = getSpPPolygonsIDSlots(NY_poly))
row.names(df) <- getSpPPolygonsIDSlots(NY_poly)

NY_polyDF = SpatialPolygonsDataFrame(NY_poly, data= df)
boundaries <- NY_polyDF


```




```{r, message=FALSE, error=FALSE, include=FALSE}

# uber_raw_2014 <- bind_rows(uber_raw_data_apr14)

uber_raw_2014 <- bind_rows(uber_raw_data_apr14, uber_raw_data_may14,     uber_raw_data_jun14, uber_raw_data_jul14, uber_raw_data_aug14, uber_raw_data_sep14)

rm(uber_raw_data_apr14, uber_raw_data_may14,     uber_raw_data_jun14, uber_raw_data_jul14, uber_raw_data_aug14, uber_raw_data_sep14)
gc() #memory cleaning

uber_raw_2014 <- uber_raw_2014 %>% mutate(
    `Date/Time` = mdy_hms(`Date/Time`))

coord <- uber_raw_2014  
coordinates(coord) = ~ Lat + Lon
proj4string(coord) <- proj4string(boundaries)
# proj4string(coord) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
coord <- spTransform(coord, proj4string(boundaries))
    
coord_over_neibourhood <- over(coord, boundaries)

uber_raw_2014_neigh <- bind_cols(uber_raw_2014, coord_over_neibourhood) 


```
```{r}
uber_raw_2014
```
```{r}
uber_raw_2014_neigh
```


```{r,  message=FALSE, error=FALSE, include=FALSE}
uber_raw_2014 <- uber_raw_2014_neigh%>%
    rename(Borough = id) %>% 
    filter(!is.na(Borough)) 

rm(uber_raw_2014_neigh, uber_2014_out, coord_over_neibourhood)
gc()
```

```{r}
uber_raw_2014
```

```{r Coordinates and Projections, include=FALSE}

uber_utm <- uber_raw_2014 

coordinates(uber_utm) <- c("Lon", "Lat")
proj4string(uber_utm) <- CRS("+proj=longlat +datum=WGS84")  ## for example

res <- spTransform(uber_utm, CRS("+proj=utm +zone=18 ellps=WGS84"))
```
```{r}
head(res)
```
```{r}
head(uber_utm)
```
```{r}
head(uber_raw_2014)
```

###  Trick to Cluster Data

When trying to cluster geo data, the fist idea may be to use Hierarchical Clustering.  To do this you have to make a distance matrix first.  Because 2014 had more the 4 million data points, we are talking about a distance matrix of 4 million by 4 million rows!!!! I don't have that much memory.

So instead what I did was to project the coordinates to UTM (in meters) and then cluster data in a 50x50 meters blocks.  Then tally how the points inside a block.

50 by 50 meters  is about (164 feet by 164 feet) that's about a quarter of the length of a New York block (the long side, from avenue to avenue).  I think is a good cluster size.

```{r Uber_UTM, include=FALSE}
uber_utm <- data.frame(uber_raw_2014$Lon, uber_raw_2014$Lat,
                            res$Lon, res$Lat, 
                            res$`Date/Time`, 
                            res$Borough) %>% 
    mutate(Lon = uber_raw_2014.Lon,
           Lat = uber_raw_2014.Lat,
           Lon_utm = res.Lon, 
           Lat_utm = res.Lat,
           `Date/Time` = res..Date.Time. ,
           Borough = res.Borough) %>% 
    select(Lon:Borough) %>% 
    mutate(Lon50m = Lon_utm %/% 50,
           Lat50m = Lat_utm %/% 50) %>% 
    group_by(Lon50m, Lat50m) %>% 
    mutate(Lon = mean(Lon),
           Lat = mean(Lat),
           Lon_utm = mean(Lon_utm),
           Lat_utm = mean(Lat_utm)) %>% ungroup() %>% 
    mutate(
        Day_of_Month = day(`Date/Time`),
        Month = month(`Date/Time`),
        Year = year(`Date/Time`),
        Hour = hour(`Date/Time`),
        Weekday = wday(`Date/Time`, label = TRUE)) %>%
    mutate(Day_Segment = Hour %/% 3 * 3) %>%
    mutate(Day_Segment = if_else(Day_Segment == 0, "00-03",
                   if_else(Day_Segment == 3, "03-06", 
                           if_else(Day_Segment == 6, "06-09",
                                   if_else(Day_Segment == 9, "09-12",
                                           if_else(Day_Segment == 12, "12-15",
            if_else(Day_Segment == 15, "15-18",
                    if_else(Day_Segment == 18, "18-21", "21-24")))))))) 


```
```{r}
head(uber_utm)
nrow(uber_utm)
```
```{r}
nrow(uber_utm$Lon)
nrow(uber_utm$Lat)
nrow(uber_utm$Weekday)
nrow(uber_utm$Day_Segment)
uber_utmprova = cbind.data.frame(uber_utm$Lon,uber_utm$Lat,uber_utm$Weekday,uber_utm$Day_Segment)
head(uber_utmprova)
```


### Party & After Party Data Frames

I'm going to make a couple of assumptions

1. The party Weekend Extends from Thurday to Sunday
2. People leave party places between 12am and 3 am (so actually, party Weekdays are Fri-Sun)
3. People leave after parties between 3am and 6am Thurday to Saturday (actually, Fri-Sun)

```{r}
colnames(uber_utmprova) = c("Lon","Lat","Weekday","Day_Segment")

Party_place <- uber_utm %>% 
    filter( Day_Segment %in% c("00-03"),
            Weekday %in% c ("sab", "dom", "lun")) %>% 
    group_by(Lon, Lat) %>% 
    tally() %>% 
    arrange(desc(n))

nrow(Party_place)
```

## Party

There were 3 main party spots in New York.  The first one was the Meat Packing District / Chelsea, the second one was East Village/Lower East Side and the third one is in Williamsburg.

I'm going to zoom in on those 3 Neighborhoods, and mark the pickup spots.  Maybe is useful for someone who lives or travels frequently to NYC

```{r Party Maps, error=FALSE, message=FALSE}




plot_party <- Party_place %>% 
    head(30) %>% 
    leaflet() %>%
    addTiles %>%
    addProviderTiles("Hydda.Full") %>%
    setView(lng = -73.98, lat= 40.733, zoom = 13) %>% 
    addMarkers()  
plot_party
```

### Party - Chelsea and Meatpacking District

There is a big cluster of pickup spots between 12th St to 14th St and from Hudson to Washington St.

1. 16th St and 9th Av
2. Both Corners of 10Av and 17 and 18 St
3. Jane St and West St
4. 14th St and 8th Av

```{r Party Chelsea, error=FALSE, message=FALSE, echo=FALSE}
plot_party <- Party_place %>% 
    head(30) %>% 
    leaflet() %>%
    addTiles %>%
    setView(lng = -74.004, lat= 40.7415, zoom = 15) %>% 
    addMarkers()  
plot_party
```


### Party - East Village and Lower East Side

There are 2 well defined clusters

1. Between Houston and Delancy Street and between Bowey and Ludlow St
2. Between Broadway and Lafayette and betwen Bond and 4th Street

```{r Party Downtown, error=FALSE, message=FALSE, echo=FALSE}
plot_downtown <- Party_place %>% 
    head(30) %>% 
    leaflet() %>%
    addTiles() %>%
    addProviderTiles("Hydda.Full") %>%
    setView(lng = -73.993, lat= 40.724, zoom = 15) %>% 
    addMarkers()  
plot_downtown
```


### Williamsburg

```{r Party Brooklyn, error=FALSE, message=FALSE, echo=FALSE}
plot_brooklyn <- Party_place %>% 
    head(30) %>% 
    leaflet() %>%
    addTiles() %>%
    addProviderTiles("Hydda.Full") %>%
    setView(lng = -73.957, lat= 40.722, zoom = 15) %>% 
    addMarkers()  
plot_brooklyn
```